Do we need detailed demographic data to forecast plant population responses to climate change?

### 2015-06-18 manuscript/manuscript.Rmd
updates ms discussion

Do we need detailed demographic data to forecast the impacts of climate change on populations?
========================================================================================

@@

Materials and Methods
---------------------
### Study site and data
Our demographic data comes from the Fort Keogh Livestock and Range Research Laboratory in eastern Montana's northern mixed prairie near Miles City, Montana, USA ($46^{\circ}$ 19' N, $105^{\circ}$ 48' W). The dataset is freely available on Ecological Archives[^link] [@Anderson2011] , and interested readers should refer to the metadata therein for a complete description. The site is about 800 m above sea level and mean annual precipitation (1878-2009) is 334 mm, with most annual precipitation falling from April through September. The site is grass dominated and, for the purposes of our study, we focus on the four most abundant graminoid species: \emph{Bouteloua gracilis} (BOGR), \emph{Hesperostipa comata} (HECO), \emph{Pascopyrum smithii} (PASM), and \emph{Poa secunda} (POSE).

@@

In this paper, we model populations based on two levels of data: individual and quadrat (Fig. 1). The individual data is the "raw" data. For the quadrat level we data we simply sum individual areal cover for each quadrat by species. This is equivalent to a perfect census of quadrat percent cover, so we do not need to consider measurement error. Based on these two datasets we can compare population models built using individual level data and aggregated quadrat level data.

@@

Equilibrium cover from both models was sensitive to climate (Fig. 2a-d). The IPM projected percent changes in equilibrium cover from `r round(min(subset(meanparam_agg, species=="BOGR" & model=="IPM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="BOGR" & model=="IPM")["meandiff"])*100)`% for *B. gracilis*, `r round(min(subset(meanparam_agg, species=="HECO" & model=="IPM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="HECO" & model=="IPM")["meandiff"])*100)`% for *H. comata*, `r round(min(subset(meanparam_agg, species=="PASM" & model=="IPM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="PASM" & model=="IPM")["meandiff"])*100)`% for *P. smithii*, and `r round(min(subset(meanparam_agg, species=="POSE" & model=="IPM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="POSE" & model=="IPM")["meandiff"])*100)`% for *P. secunda*. The QBM projected opposite and greater percent changes in equilibrium cover for *B. gracilis* (`r round(min(subset(meanparam_agg, species=="BOGR" & model=="QBM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="BOGR" & model=="QBM")["meandiff"])*100)`%) and *H. comata* (`r round(min(subset(meanparam_agg, species=="HECO" & model=="QBM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="HECO" & model=="QBM")["meandiff"])*100)`%; Fig. 2a-b). For *P. smithii*, the QBM projected opposite changes in equilibrium cover than the IPM, but of similar magnitude (`r round(min(subset(meanparam_agg, species=="PASM" & model=="QBM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="PASM" & model=="QBM")["meandiff"])*100)`%; Fig. 2c). *P. secunda* was the only species that the IPM and QBM made projections of the same sign and somewhat similar magnitude (Fig. 2d). As expected based on model validation (Table 1), IPM projections were more uncertain than QBM projections for all species and all climate change scenarios (Fig. 2e-h).

The response of a population to climate change is a result of the aggregate effects of climate on individual vital rates. Since the IPM approach relies on vital rate regressions, we were able to quantify the sensitivity of each vital rate in isolation and in pairwise combinations. Species showed similar trends (Fig. 3). Growth was the most sensitive vital rate for all species, showing a negative response to increased precipitation, and stronger positive response to increased temperature, and a mostly positive response when both climate factors are increased (Fig. 3). *B. gracilis* survival rates were sensitive to temperature, resulting in an increase in plant cover under increased temperature (Fig. 3a). In isolation, recruitment and survival were insensitive to climate factors for *H. comata* (Fig. 3b). Survival and recruitment of *P. smithii* were both sensitive, negatively, to temperature and precipitation (Fig. 3c). *P. secunda* equilibrium cover was sensitive to the climate effects on survival and recruitment, showing a negative effect on both vital rates for increased precipition, but a strong positive effect on survival with increased temperature (Fig. 3d). The climate impact of recruitment on equilibrium cover was negative for precipitation and temperature increases (Fig. 3d).

@@

As a potential remedy to the "data dearth" problem, @Queenborough2011 and @Freckleton2011, building on work by @Taylor2004, advocate a "density-structured" modeling approach. Such models do not require individual level demographic data. Indeed, @Queenborough2011 provide empirical evidence that density-structured models adequately described population dynamics. The results from density-structured models are not as precise as those from traditional population models, but the loss in precision is traded off with a gain in data. The study by @Queenborough2011 included data from 500 fields (4 hectares each) in 49 farms, all collected by two people in 6 weeks. This is far more data from a far greater spatial extent than possible if measuring individual plant demography (in a world of limited time and money, at least). The appeal of density-structured approaches is clear.

@@

Indeed, when we perturbed climate factors the QBM made forecasts completely contradictory to those of the IPM for three of our four study species (Fig. 2a-d). In a perfect world, the QBM would have made forecasts of at least the same direction as the IPM. If that had been the case we could conclude that aggregate level models could prove useful for forecasting climate change impacts on populations. Unfortunately, this was not the case.

@@

This result is not as dire as it seems when we consider model uncerainty. Yes, the QBM failed to match IPM forecasts under perturbed climate, but the uncertainties around those forecasts were so great (Fig. 2e-h) that we cannot say that the two model's forecasts were different with any level of confidence. This offers hope that with "enough" population level data one will be able to reduce parameter uncertainty and thus model forecasts. This remains an open question.

### The role of vital rate climate dependence
Our intuition was that species with strong climate-dependence on vital rates not well resolved at the aggregate level would result in different forecasts from the two models. For example, survival is very size dependent: smaller individuals have a higher probability of death [@Chu2014]. At the same time, a single small individual contributes relatively little to percent cover estimates at the plot scale. So, if survival of individuals was positively impacted by temperature increases, for example, we would expect to detect this signal in the individual level data but not in the aggregate level data.

### Forecasting the future, and the future of forecasting
Our goal was not make any explicit forecast for the future state of these populations based on predicted climate change. But our results highlight the state of affairs in ecology when it comes to forecasting the impacts of climate change. The analysis we conducted here could be considered, with some exceptions of course, at the forefront of ecological forecasting in terms of the statistical approach employed (hierarchical Bayesian), the type of population model we used (stochastic IPM with parameter uncertainty), and the amount of data we had at our disposal (14 years of individual-level data). Yet, model predictions proved so uncertain that any forecast, when bounded with uncertainty, would be at best not useful and at worst meaningless.

Something about fitting the models...then cite Britta's paper: 20-25 years needed!

@@

Figures
-------
```{r figure_1,dependson="plot-options", fig.cap="Work flow of the data aggregation, model fitting, and population simulating.", fig.height=6}
library(png)
library(grid)
img <- readPNG("components/figure/micromeso_flowchart.png")
grid.raster(img)
```

@@

```{r figure_2, dependson="plot-options", fig.cap="Proportional change in species' mean cover caused by a 1% increase in observed precipitation (+ppt), temperature (+temp), or both (+ppt&temp) as predicted by the individual-based IPM and the aggregate-based QBM. Top panels show the mean predicted proportional change in cover; lower panels show the 90% quantiles of predicted proportional cover change. Mean predictions and their associated uncertainties are shown in separate panels for visual clarity.", fig.width=8.5, fig.height=6, cache=FALSE, results='hide'}
setwd("~/Repos/MicroMesoForecast/analysis/")

@@

dgd = position_dodge(width = 0.6)
uncertainplot <- ggplot(meanparam_agg)+
  geom_errorbar(aes(x=climsim, ymax=hidiff, ymin=lodiff, group=model, linetype=model),
                position=dgd, width=0.25)+
  theme_bw()+
  scale_linetype_discrete(name="Model")+
  facet_wrap("species", nrow=1, scales="free")+
  scale_x_discrete(labels=c("+ppt", "+temp", "+ppt&temp"))+
  ylab("Proportional difference")+
  xlab("Climate simulation")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

@@

```{r figure_3, dependson="plot-options", fig.cap="Sensitivity of equilibrium cover to a 1% increase in precipitation (+ppt), temperature (+temp), or both (+ppt&temp) applied to individual and combined vital rates. For example, the points associated with G show the median cover from IPM simulations where a climate perturbation is applied only to the growth regression climate covariates.", fig.width=4.5, fig.height=6, cache=FALSE, results='hide'}
setwd("../analysis/ipm/simulations/results/climate_sensitivity/")
files <- list.files()
cover_files <- files[grep("cover", files)]
spp_id <- substr(cover_files, 1, 4)
num_files <- length(cover_files)
all_sims <- data.frame(time=NA, cover=NA, species=NA, climsim=NA, vital=NA)
for(i in 1:num_files){
  tmp <- read.csv(cover_files[i])
  tmp$species <- rep(spp_id[i], nrow(tmp))
  all_sims <- rbind(all_sims, tmp)
}
all_sims <- all_sims[2:nrow(all_sims),]

>>>>

Do we need detailed demographic data to forecast the impacts of climate change on plant populations?
========================================================================================

@@

Materials and Methods
---------------------
### Study site and data
Our demographic data comes from the Fort Keogh Livestock and Range Research Laboratory in eastern Montana's northern mixed prairie near Miles City, Montana, USA ($46^{\circ}$ 19' N, $105^{\circ}$ 48' W). The dataset is freely available on Ecological Archives[^link] [@Anderson2011] , and interested readers should refer to the metadata therein for a complete description. The site is about 800 m above sea level and mean annual precipitation (1878-2009) is 334 mm, with most annual precipitation falling from April through September. The site is grass dominated and, for the purposes of our study, we focus on the four most abundant graminoid species: \emph{Bouteloua gracilis} (BOGR), \emph{Hesperostipa comata} (HECO), \emph{Pascopyrum smithii} (PASM), and \emph{Poa secunda} (POSE) (Fig. 1).

@@

In this paper, we model populations based on two levels of data: individual and quadrat (Fig. 2). The individual data is the "raw" data. For the quadrat level we data we simply sum individual areal cover for each quadrat by species. This is equivalent to a perfect census of quadrat percent cover, so we do not need to consider measurement error. Based on these two datasets we can compare population models built using individual level data and aggregated quadrat level data.

@@

Equilibrium cover from both models was sensitive to climate (Fig. 3a-d). The IPM projected percent changes in equilibrium cover from `r round(min(subset(meanparam_agg, species=="BOGR" & model=="IPM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="BOGR" & model=="IPM")["meandiff"])*100)`% for *B. gracilis*, `r round(min(subset(meanparam_agg, species=="HECO" & model=="IPM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="HECO" & model=="IPM")["meandiff"])*100)`% for *H. comata*, `r round(min(subset(meanparam_agg, species=="PASM" & model=="IPM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="PASM" & model=="IPM")["meandiff"])*100)`% for *P. smithii*, and `r round(min(subset(meanparam_agg, species=="POSE" & model=="IPM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="POSE" & model=="IPM")["meandiff"])*100)`% for *P. secunda*. The QBM projected opposite and greater percent changes in equilibrium cover for *B. gracilis* (`r round(min(subset(meanparam_agg, species=="BOGR" & model=="QBM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="BOGR" & model=="QBM")["meandiff"])*100)`%) and *H. comata* (`r round(min(subset(meanparam_agg, species=="HECO" & model=="QBM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="HECO" & model=="QBM")["meandiff"])*100)`%; Fig. 3a-b). For *P. smithii*, the QBM projected opposite changes in equilibrium cover than the IPM, but of similar magnitude (`r round(min(subset(meanparam_agg, species=="PASM" & model=="QBM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="PASM" & model=="QBM")["meandiff"])*100)`%; Fig. 3c). *P. secunda* was the only species that the IPM and QBM made projections of the same sign and somewhat similar magnitude (Fig. 3d). As expected based on model validation (Table 1), QBM projections were more uncertain than IPM projections for all species except *P. smithiii* (Fig. 3e-h).

The response of a population to climate change is a result of the aggregate effects of climate on individual vital rates. Since the IPM approach relies on vital rate regressions, we were able to quantify the sensitivity of each vital rate in isolation and in pairwise combinations. Species showed similar trends (Fig. 4). Growth was the most sensitive vital rate for all species, showing a negative response to increased precipitation, and stronger positive response to increased temperature, and a mostly positive response when both climate factors are increased (Fig. 4). *B. gracilis* survival rates were sensitive to temperature, resulting in an increase in plant cover under increased temperature (Fig. 4a). In isolation, recruitment and survival were insensitive to climate factors for *H. comata* (Fig. 4b). Survival and recruitment of *P. smithii* were both sensitive, negatively, to temperature and precipitation (Fig. 4c). *P. secunda* equilibrium cover was sensitive to the climate effects on survival and recruitment, showing a negative effect on both vital rates for increased precipition, but a strong positive effect on survival with increased temperature (Fig. 4d). The climate impact of recruitment on equilibrium cover was negative for precipitation and temperature increases (Fig. 4d).

@@

As a potential remedy to the "data dearth" problem, @Queenborough2011 and @Freckleton2011, building on work by @Taylor2004, advocate a "density-structured" modeling approach. Such models do not require individual level demographic data and can adequately describe population dynamics [@Queenborough2011]. The results from density-structured models are not as precise as those from traditional population models, but the loss in precision is traded off with a gain in data. The study by @Queenborough2011 included data from 500 fields (4 hectares each) in 49 farms, all collected by two people in 6 weeks. This is far more data from a far greater spatial extent than possible if measuring individual plant demography (in a world of limited time and money, at least). The appeal of density-structured approaches is clear.

@@

Indeed, when we perturbed climate factors the QBM made forecasts completely contradictory to those of the IPM for three of our four study species (Fig. 3a-d). In a perfect world, the QBM would have made forecasts of at least the same direction as the IPM. If that had been the case we could conclude that aggregate level models could prove useful for forecasting climate change impacts on populations. Unfortunately, this was not the case.

@@

This result is not as dire as it seems when we consider model uncerainty. Yes, the QBM failed to match IPM forecasts under perturbed climate, but the uncertainties around those forecasts were so great (Fig. 3e-h) that we cannot say that the two model's forecasts were different with any level of confidence. This offers hope that with "enough" population level data one will be able to reduce parameter uncertainty and thus model forecasts. This remains an open question.

### The role of vital rate climate dependence
Our intuition was that species with strong climate-dependence on vital rates not well resolved at the aggregate level would result in different forecasts from the two models. For example, survival is very size dependent: smaller individuals have a higher probability of death [@Chu2014]. At the same time, a single small individual contributes relatively little to percent cover estimates at the plot scale. So, if survival of individuals was positively impacted by temperature increases, for example, we would expect to detect this signal in the individual level data but not in the aggregate level data. **PETER: do you think this is the right direction to go with this?**

### Forecasting the future, and the future of forecasting
Our goal was not make any explicit forecast for the future state of these populations based on predicted climate change. But our results highlight the state of affairs in ecology when it comes to forecasting the impacts of climate change. The analysis we conducted here could be considered, with some exceptions of course, at the forefront of ecological forecasting in terms of the statistical approach employed (hierarchical Bayesian), the type of population model we used (stochastic IPM with parameter uncertainty), and the amount of data we had at our disposal (14 years of individual-level data). Yet, model predictions proved so uncertain that any forecast, when bounded with uncertainty, would be at best not useful and at worst meaningless. How might we improve on this state of affairs?

First, forecasts could be improved by matching the spatial scale of predictor variables with the spatial scale of observations. One of the major limitations of the models we fit here is that the climate data are at a much larger scale than the plot and individual level observations of plant cover and size. Climate covariates only vary by year, with no spatial variability within years. Thus, even if we fit models to individual level data, we are missing the key interaction point between weather and individual plants. **PETER: any good citations here for decent population models fit with fine-scale climate/environmental data??**

Second, accurately detecting climate signals will take even longer time series. Recent theoretical work on detecting climate signals in noisy data suggests that even advanced approaches to parameter fitting like LASSO and random forest require 20-25 year time series (Teller et al., in review).

@@

Figures
-------
```{r figure_1, dependson="plot-options", fig.cap="Time series of average percent cover of our four focal species: *Bouteloua gracilis* (BOGR), *Hesperostipa comata* (HECO), *Pascopyrum smithii* (PASM), and *Poa secunda* (POSE).", fig.height=4}
obs_data <- read.csv("../analysis/speciesData/quadAllCover.csv")
#remove zeros for averaging
obs_data <- obs_data[which(obs_data$propCover>0),]
avgobs <- ddply(obs_data, .(year, Species), summarise,
                mean_cover = mean(propCover))
ggplot(avgobs, aes(x=(1900+year), y=mean_cover*100, shape=Species, linetype=Species))+
  geom_line()+
  geom_point(size=5, color="white")+
  geom_point(size=3)+
  ylab("Mean cover (%)")+
  xlab("Year")+
  theme_bw()
```

```{r figure_2,dependson="plot-options", fig.cap="Work flow of the data aggregation, model fitting, and population simulating.", fig.height=6}
library(png)
library(grid)
img <- readPNG("components/figure/micromeso_flowchart.png")
grid.raster(img)
```

@@

```{r figure_3, dependson="plot-options", fig.cap="Proportional change in species' mean cover caused by a 1% increase in observed precipitation (+ppt), temperature (+temp), or both (+ppt&temp) as predicted by the individual-based IPM and the aggregate-based QBM. Top panels show the mean predicted proportional change in cover; lower panels show the 90% quantiles of predicted proportional cover change. Mean predictions and their associated uncertainties are shown in separate panels for visual clarity.", fig.width=8.5, fig.height=6, cache=FALSE, results='hide'}
setwd("~/Repos/MicroMesoForecast/analysis/")

@@

####
##  Vary parameter runs
####
meanparam_ipm <- data.frame(cover=NA, species=NA, 
                             climsim=NA, model=NA)
##  IPM 
path <- "./ipm/simulations/results/climchange_varyparams/"
ipmfiles_mean <- list.files(path)
ipmfiles_mean <- ipmfiles_mean[grep("cover", ipmfiles_mean)]
for(file in ipmfiles_mean){
  tmp <- read.csv(paste(path,file,sep=""))
  tmp$model <- "IPM"
  meanparam_ipm <- rbind(meanparam_ipm, tmp[,c("cover", "species", "climsim", "model")])
}
meanparam_ipm <- meanparam_ipm[2:nrow(meanparam_ipm),]

newd <- data.frame(cover=NA, species=NA, 
                   climsim=NA, model=NA, obscover=NA)
for(dospp in unique(meanparam_ipm$species)){
  tmp <- subset(meanparam_ipm, species==dospp)
  obscover <- subset(tmp, climsim=="obs")["cover"]
  tmp2 <- subset(tmp, climsim!="obs")
  tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
  newd <- rbind(newd, tmp2)
}
ipm_meansims <- newd[2:nrow(newd),]
ipm_meansims$propdiff <- with(ipm_meansims, log(cover*100)-log(obscover*100))
# ggplot(subset(ipm_meansims, species=="PASM"))+
#   geom_histogram(aes(x=cover))+
#   facet_wrap("climsim", scales = "free")+
#   scale_x_continuous(limits=c(0,0.05))
##  QBM
meanparam_qbm <- data.frame(cover=NA, species=NA, 
                             climsim=NA, model=NA)
path <- "./quadBM/simulations/results/climatechange_varyparams/"
qbmfiles_mean <- list.files(path)
for(file in qbmfiles_mean){
  tmp <- readRDS(paste(path,file,sep=""))
  tmp$model <- "QBM"
  meanparam_qbm <- rbind(meanparam_qbm, tmp[,c("cover", "species", "climsim", "model")])
}
meanparam_qbm <- meanparam_qbm[2:nrow(meanparam_qbm),]

##  Merge IPM and QBM mean parameter results
bothmodel_outs <- rbind(ipm_meansims, qbm_meansims)
meanparam_agg <- ddply(bothmodel_outs, .(species, climsim, model), summarise,
                       meandiff = mean(propdiff),
                       hidiff = quantile(propdiff, probs = 0.95),
                       lodiff = quantile(propdiff, probs = 0.05))
dgd = position_dodge(width = 0.6)
uncertainplot <- ggplot(meanparam_agg)+
  geom_errorbar(aes(x=climsim, ymax=hidiff, ymin=lodiff, group=model, linetype=model),
                position=dgd, width=0.25)+
  theme_bw()+
  scale_linetype_discrete(name="Model")+
  facet_wrap("species", nrow=1, scales="free")+
  scale_x_discrete(labels=c("+ppt", "+temp", "+ppt&temp"))+
  ylab("Proportional difference")+
  xlab("Climate simulation")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

@@

```{r figure_4, dependson="plot-options", fig.cap="Sensitivity of equilibrium cover to a 1% increase in precipitation (+ppt), temperature (+temp), or both (+ppt&temp) applied to individual and combined vital rates. For example, the points associated with G show the median cover from IPM simulations where a climate perturbation is applied only to the growth regression climate covariates.", fig.width=4.5, fig.height=6, cache=FALSE, results='hide'}
setwd("../analysis/ipm/simulations/results/climate_sensitivity/")
files <- list.files()
cover_files <- files[grep("cover", files)]
spp_id <- substr(cover_files, 1, 4)
num_files <- length(cover_files)
all_sims <- data.frame(time=NA, cover=NA, species=NA, climsim=NA, vital=NA)
for(i in 1:num_files){
  tmp <- read.csv(cover_files[i])
  tmp$species <- rep(spp_id[i], nrow(tmp))
  all_sims <- rbind(all_sims, tmp)
}
all_sims <- all_sims[2:nrow(all_sims),]
